\documentclass[10pt]{NSF}

\begin{document}

\section{Data Management Plan}
\label{sec:data_plan}

\subsection{Types of Data}

This project will generate the following types of data:
\begin{itemize}
\item{Engineering drawings and circuit diagrams for custom printed
  circuit boards.}
\item{Field programmable gate array (FPGA) firmware blocks and
  compiled BIT files.}
\item{Monitor and control (M\&C) software and implementations of radio
  frequency interference (RFI) excision algorithms.}
\item{Laboratory- and telescope-generated high-level test data
  (e.g. digitized raw voltages and detected spectra).}
\item{A database of common sources of RFI and their characteristics.}
\item{Technical documents, memos, and publications.}
\item{Educational curricular materials.}
\item{Educational evaluation data.}
\end{itemize}

\subsection{Data Formats and Standards}

We will use common data formats wherever possible to facilitate broad
dissemination and use of our work.  This will include:
\begin{itemize}
\item{Engineering mechanical drawings and block diagrams created with
  AutoCAD and/or Autodesk Inventor.  Printable circuit board layouts
  will generate standard Gerber files that are used by manufacturing
  companies.  We will use the Eagle printable circuit board diagram
  layout program to draw schematics and design boards.}
\item{FPGA designs will generate BIT files and BOF files that are
  common in CASPER.}
\item{C++ and Python for M\&C software; Python, C++, CUDA, and Verilog
  HDL and Simulink for RFI excision algorithms.}
\item{High-level test data will primarily be in the FITS format.
  SDFITS is an implementation for spectral line data supported by GBO;
  PSRFITS is an implementation for pulsar data supported by numerous
  third-party software packages.  Voltage data will in the GUPPI RAW
  format, which is used for other CASPER-based back-ends and supported
  by numerous third-party software packages.}
\item{SQL will be used for the RFI database.}
\item{Documents, including educational materials, will be written in
  portable formats such as PDF.}
\item{Educational evaluation data will be stored using spreadsheets.}
\end{itemize}

\subsection{Usage and Distribution Policies}

A public-facing project web page will be hosted by GBO for
communicating results of our work to a scientific, technical, and
general audience.  It will be updated with status reports upon the
completion of key milestones, and will include links to
publicly-available data products.  An internal wiki page will be
hosted by GBO for recording meeting minutes, internal project memos,
and preliminary data products.
\begin{itemize}
\item{Engineering mechanical drawings and circuit board layouts will
  be placed on GBO GitHub repositories.}
\item{We will distribute all resources that enable the integration of
  new hardware into the CASPER tool-flow through the official CASPER
  GitHub repository}
\item{Source code and firmware for RFI excision algorithms will be
  made placed on GBO GitHub repositories.}
\item{Test-data are may be many tens to hundreds of gigabytes in size,
  and will be hosted on spinning-disk at GBO.  Data generated during
  initial testing and debugging will be retained for use by the
  project team.  Final validation and verification data will be made
  publicly available.  High-level descriptions and meta-data of final
  validation and verification data will be published on the public
  project web page.  Final validation and verification data sets will
  be made available via FTP.}
\item{The RFI database will be publicly accessible via a GBO-hosted
  web page.  Community contributions will be welcome with editing
  privileges granted on a case-by-case basis via coordination with the
  PI and the GBO computing division.}
\item{Final commissioning reports, technical publications, and
  curricular material will be made freely available via the public
  project web page.}
\item{Since evaluation data will not be identified with particular
  individuals, institutional review board approval is not required and
  the pre- and post-data will be available upon request by GBO.}
\end{itemize}

\subsection{Data Archiving}

GBO will curate and archive data.  Documentation, firmware blocks and
BIT files, test data, public databases, and public and internal
web pages will be stored locally on spinning disks that undergo regular
back-up as part of normal GBO computing division activities.  Local
revision control repositories will be used to track changes in
firmware and software.

\end{document}
